Comparing vector-based and ACT-R memory models of tag retrieval:\protect\\User-customized hashtag/tag prediction on Twitter and StackOverflow

### 2014-07-22 writeups/dissertation/dissertation.tex
foo

The same StackOverflow popular-users datasets used to test the prior component in isolation were used to test the prior component when context was added.
However, since computing activations for the context component take an order of magnitude more wall-clock time than computing the prior comonent, not all observations could be included for the test.
To reduce the observations, the last FIXME posts from each user in the datasets were sampled.
Using the last observations for each user ensures that the observations tested have aquired the greatest amount of past tagging history in the datasets.
So if a context-only model performs just as well as the full model in this case, it would be surprising.

@@

\subsubsection{Method}
\subsubsection{Results}

@@

FIXME: Include a paragraph about how the random-permutaiton model works so well due to the compression abilities, and not because of word order.

\section{Summary}

This research has numerous components that are relevant to improving our understanding of declarative memory.
It should be emphasized that it has only recently become possible to evaluate psychological theories of declarative memory on large-scale real-world tasks where the theory can be tested on hundreds of millions of data points.
These information-rich large-scale environments provide a unique opportunity to really stress test and explore the impacts of the psychological constraints of each theory
on a much larger scale than has previously been possible.
This environment will allow, for example, overall performance differences between an ACT-R Bayesian retrieval model and a random-permutation vector-based model to be directly compared on two real-world tagging tasks,
which has not been done previously.

This data-rich environment will also enable the proposed modifications to each model to be quickly evaluated, so that further model improvements can be iteratively proposed and tested.
The main two proposed model modifications (incorporating word order into ACT-R and user history into vector-based representations) have not been tested before, and seem likely to significantly improve performance.

Further, the proposed model modifications are aimed at the architectural level, may very well be task independent, and may generalize and improve model fit across other tasks.
Discovering an effective way to represent the time decay of information in a vector-based model could improve model fit on other tasks where the memory model has been used. 
Finding a task-agnostic way to represent word order information in a Bayesian ACT-R model could improve model accuracy when applied to other domains where it seems plausible that participants are utilizing word order.

Additionally, the tag-recommendation models developed through this research should be immediately relevant to real-world tasks.
If model accuracy is high enough, then the developed models may be used as a foundation for tag recommendation systems that are deployed on Twitter or StackOverflow.

Also since the developed models will be task agnostic, they can potentially be used as tag recommendation systems for other domains.
Examples of similar tagging tasks are hashtag use on Facebook, flagging (i.e., tagging) Wikipedia articles for content issues or overly-opinionated material,
auto labeling incoming mail and placing it into appropriate folders, and spam filtering.
Having an accurate and efficient co-occurrence memory model that correctly retrieves relevant tags can be applied to a wide variety of real-world tasks.

The end result of this research will be [1] a better understanding of how a user's past tagging history influences future tag use,
[2] a more thorough evaluation of the strengths, weaknesses, and differences between two cognitively-plausible memory models on large-scale real-world retrieval tasks,
[3] improvements to each memory model where the strengths of each model (word order and user prior likelihood) are incorporated into the other,
and [4] two efficient implementations of the models for making user-customized tag predictions for StackOverflow posts and Twitter tweets. 
Introducing, exploring, and evaluating reasonable architectural modifications to two commonly-used memory models is worth the effort:
It will increase our understanding of how the effects of imposing specific architectural constraints on memory models influence what information is retrieved.

>>>>

The same StackOverflow popular-users datasets used to test the prior component in isolation were used to test the prior component when context was added.
However, since computing activations for the context component take an order of magnitude more wall-clock time than computing the prior comonent, not all observations could be included for the test.
To reduce the observations, the most recent FIXME posts from each user in the datasets were sampled.
Using the most recent observations for each user ensures that the observations tested have aquired the greatest amount of past tagging history in the datasets.
So if a context-only model performs just as well as the full model in this case, it would be surprising.

@@

The co-occurrence matrix for the context component of both models has been built by aggregating across users and collecting all instances of context words and associated tags.
That is, the default co-occurrence matrix is not customized to each user, so the same matrix is used for all users.
This is commonly done mainly because there is never enough context by tag co-occurrences for a single user to generate a stable co-occurrence matrix.
If there was enough information, acuracy would most likely improve since it is not likely that every user's word by tag co-occurrence profile is exactly the same.
However, since several hundred posts for each user in the popular-users datasets are available, an analysis was tried where user-custom co-occurence matrices were used.

Separate co-occurrence matrices were built for each user in the sampled StackOverflow and Twitter popular-users dataset.
The same sampling method used to test the impact of the prior component compared to the context component for the popular-users datasets was used,
but the oldest posts that were not included in the testing set were used to build the co-occurrence matrices.
So for each user, the co-occurrence matrix was created from their oldest posts, up to (and not including) their FIXME newer posts that were used to test model accuracy.

\subsubsection{Result}

Model accuracy when using a user-customized co-occurrence matrix for StackOverflow and Twitter is shown in Figures \ref{figPUserCustomSO} and \ref{figPUserCustomT}.

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-userSjiSO-crop.pdf}}}
  \caption{Model accuracy with user-customized context compoenent for StackOverflow}
  \label{figPUCustomSO}
\end{figure}

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-userSjiT-crop.pdf}}}
  \caption{Model accuracy with user-customized context component for Twitter}
  \label{figPUCustomT}
\end{figure}

The results compare accuracy for models that use a global co-occurrence and user-specific co-occurence matrix.
For StackOverflow, model accuracy for the user-specific (labeled as ``user sji'') can be compared against the default global co-occurrence matrix.
Accuracy decreases dramatically when a user-specific co-occurrence matrix is used.

The results are similar for Twitter.
However since the entire dataset is not available for Twitter, a global co-occurrence matrix condition can not be tested,
since the popular-users dataset samples across a wide range of hashtags, and due to Twitter's API restrictions it would be infeasable to collect all posts for each of the hashtags that are used in the dataset.
But user-specific co-occurence matrices can be built, since the popular-users dataset does contain all context and hashtag observations for each user.

Model accuracy when only the single user-customized context predictor is used is relatively poor, and the prior term is a much better predictor.
Further, when this component is combined with the prior component, model performance improves only slightly over performance when only the prior is used.

Both models require large numbers of context co-occurrence observations to make stable predictinos from context information.
Otherwise, the co-occurrence matrices do not have enough data to properly stabilize.
Millions of posts were used to generate the global co-occurrence matrices for Twitter and StackOverflow.
If this number of posts had been collected for a single user, then the user-customized co-occurrence matrix for that user should lead to more accurate predictions compared to using the global matrix.
However, only around 100 posts were used to build the user-specific matrices for each user.

This is most likly why model accuracy was so poor when user-specific matrices were used:
Even for the large datasets analyzed in this study, there is still not enough history collected for each user to build customized co-occurrence matrices.
The prior model component can be customized for each user, since the prior is only concerned with tag use, and there are a reasonbly small set of tags that the model is choosing from.
However, the context component is built from the the set of all of the prior tags crossed with all of the possible words, which is a much larger space than just the set of tags.
So building a stable context component requires a much larger set of data than the prior component.

That said, if a user only chooses a small set of tags (so the tag space is small), and computing a global co-occurrence matrix is infeasable (e.g., with the Twitter popular-users dataset), 
then it may be worthwhiel to add the user-custom context model compoennt as an additional predictor.
As was shown for Twitter, model accuracy can slightly improve over a model that just uses a custom prior, even when the co-occurrence matrix only contains around 100 posts.

@@

\subsubsection{Representation}

Overall, both the random-permutaiton and Bayesian models fit the observed tagging behavior in StackOverflow and Twitter relatively well.
Given that the storage space required to represent the co-occurrences for the random-permutation is dramatically smaller than the uncompressed Bayesian model,
and that the compression is lossy and essentially random (i.e., no SVD),
it is quite interesting that the random-permutation model performs nearly as well as the full Bayesian model for this task.

These results pose an interesting and important fundamental question about declarative memory:
How, precisely, is the neural system configured to represent co-occurrences for long-term memory?
Since, at the behavioral level, both models perform similarly, it is possible that the neural system is using some sort of vector-based compressed storage (e.g., random permutations) to store co-occurrences.
Vector-based models are simple enough that they can easily be implemented in a neural network.
Perhaps a form of vector-based compressed storage is being used at the neural level,
since it can be implemented in neurons and it is a way for the machinery to generate predictions that closely match the statistical structure of the environment (i.e., that closely match the Bayesian representation).

As neural imaging systems improve, this question of neural representation can be best answered by looking directly at the neural activity.
However, there are possible behavioral studies that could be done to provide evidence that may favor one representation over the other:
The largest differences between the Bayesian and vector-based models happen with small numbers of co-occurrence observations.
The vector-based model begins to generate more accurate predictions sooner than the Bayesian models,
since the noise introduced from the compression helps smooth out activation values until a larger number of co-occurrences are observed and the co-occurrence values stabilize.
For a behavioral study, novel words could be introduced alongside the corpus of known words, and the participant could be asked to make retrievals of the novel words and known words as more novel words are introduced.
Both models would make different predictions about the rate at which activation values for the novel words stabilize relative to the known words, and such make different predictions about retrieval error rates over time.
Learning a new language might be a domain where large amounts of these sorts of retrievals could be collected.
For instance, the online site Duolingo might contain enough collective data about language learning and word translation to better tease apart the two context models.

\subsubsection{Compression}

Also, the relatively strong performance of the random-permutation model is not due to the model's ability to represent word order.
Performance is similar for the random-permutation and Bayesian models when the co-occurrence matrix for both models is the same, and not based on word order.
The random-permutation model seems to work so well primarily due to the way that the full co-occurrence representation is compressed.

Compression for the random-permutation model is essentiallly random.
That is, no SVD is done to identify the most predictive components,
so the compressed representation used by the random-permutation model does not contain nearly as much of the variance from the original co-occurrence matrix that it could have.
However, performing a full SVD on a co-occurrence matrix as large as the ones generated for StackOverflow and Twitter is computationally infeasable.
So, it also seems plausible that this SVD computation is infeasable at the neural level.
\textcite{Landauer1997} explicitly state that they make no claims that the brain is computing an SVD at the neural level;
only that the neural machinery is set up so that the representation makes predictions that are similar to what would have been generated if the matrix decomposition for an SVD would have been computed.
Perhaps the randomly-compressed representation used for the random-permutaiton model is the way in which the brain tries to achive the full LSA- and SVD-like approach that \citeauthor{Landauer1997} propose.

\subsubsection{Attentional Weight}

The declarative memory model for ACT-R has an attentional weight term $W_{j}$ that can be used to implement how stop words are handled.
The random-permutaiton model was also easily modified to have an attentional weight term by attentuating each word's set of ones and negative ones values by that word's attentional weight.
Both a method for stop-word removal (frequency filtering) and stop-word attentuating (entropy weighting) performed well for both models.
Also, using a proper method for handling the low-predictor words produced some of the largest improvements in model accuracy compared to the other explorations
(e.g., compared to word order, properly adding terms, compression amount for random-permutation representation).
Since the ACT-R declarative memory theory already has a component for attentional weight, it may be the case that the neural system performs some form of stop-word weighting when cmoputing activations,
and this is reflected in the attentional weight term.
That is, $W_{j}$ can be defined mathematically, has the same general form across tasks, and both the frequency-filtering and entropy-weighting methods for stop-word removal are likely first candidates for the definition.
This claim, however, would require that no top-down activatino is used to compute the attentional weight, which would mean that the $W_{j}$ is not under conscious control, which is not likely to be correct.
Perhaps it is a combination of top-down and bottom-up systems that determine $W_{j}$, and the entropy-weighting and frequency-filtering techniques are methods of formalizing the bottom-up component.

\subsubsection{Generalizing Results}

It seems likely that the entropy method for computing the attentional weight term $W_{j}$ may be applicable to other declarative memory tasks.
The entropy method is parameter free (unlike the frequency-filtering method), so it completely data driven (much like the co-occurrence matrix) and does not require tuning.
It is common to handle stop words in some manner when working with NLP and large text documents.
So, if these large natural-language documents are used to test declarative memory, then the entropy method may be a more cognitively plausible way of dealing with stop words rather than 
filtering based on a pre-determined list or based on the frequency count in the dataset, particulary since the entropy method substantially increases model accuracy for these datasets and since it is parameter free.

Adding a prior component to the random-permutatino model also increased model accuracy for both StackOverflow and Twitter.
It is unsurprising that this occurred.
However, it is quite surprising that this is the first time that the prior component has been included in a random-permutation model.
If the random-permutatino model is explored as a declarative memory theory for other datasets, it should be worthwhile to include the prior component for those explorations.
Although it did not seem to matter much how the prior component is combined with the context component in this study, 
there may be better ways than the log-odds transformation to combine the terms.
Further research could focus on exploring a larger set of methods for combining the context component of the random-permutation model with the prior component from the ACT-R theory.

\subsubsection{Future Applications}

The tag-recommendation models developed through this research should be immediately relevant to real-world tasks.
Model accuracy is high enough (particularly for StackOverflow) for the developed models to be used as a foundation for tag recommendation systems that are deployed on Twitter or StackOverflow.
Also since the developed models are task agnostic, they can potentially be used as tag recommendation systems for other domains.
Examples of similar tagging tasks are hashtag use on Facebook, flagging (i.e., tagging) Wikipedia articles for content issues or overly-opinionated material,
auto labeling incoming mail and placing it into appropriate folders, and spam filtering.
Having an accurate and efficient co-occurrence memory model that correctly retrieves relevant tags can be applied to a wide variety of real-world tasks.

\subsubsection{Summary}

The end result of this research is [1] a better understanding of how a user's past tagging history influences future tag use,
[2] a more thorough evaluation of the strengths, weaknesses, and differences between two cognitively-plausible memory models on large-scale real-world retrieval tasks,
and [3] two efficient implementations of the models for making user-customized tag predictions for StackOverflow posts and Twitter tweets. 
This work has shown how past user history is important, and how a context-only model like the random-permutation model can be modified to include the prior component.
The work has also shown how the strength of the random-permutation model is not in its ability to represent word order,
and rather in its ability to efficiently compress the full co-occurrence representation used for the Bayesian model.
Several different formalisms for ACT-R's attentional weighting term were also explored, and an entropy-weighting technique was found to perform well without requiring any parameter tuning.
Finally, this work has helped raise the question about how co-occurrence observations are represented at the neural level, since two models with very different representations performed similarly well.
The results from the study and the new research questions raised will help to better understand how imposing specific architectural constraints on memory models influence what information is retrieved.

\subsubsection{Acknowledgements}

The majority of the modeling and analysis was done with the R statistical computing environment \parencite{RCoreTeam2014}, and the data.table \parencite{Dowle2014} and ggplot2 \parencite{Wickham2009} packages.
